---
title: "ggplot_example_code"
author: "Hannah Pook"
date: '2022-05-12'
output: html_document
---
Load in packages which are needed 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise 1

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.
The data-set being used is 'diamonds' data-set
Can find out more info using `?diamonds` function 

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer: creates a grey block, this is because we haven't said what we want on the x axis or y axis, or any other information

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer: this command defines the x and y axis. Note that the range is not random (it is based on the points in the data-set). However, we haven't said how we want our datat to be plotted - we will add this now.

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
  
```

- Color data points by `cut`.

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour=cut))+
  geom_point()
  
```

This has created a plot in which points are coloured by cut

- Add a smoothed mean trend line.

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour=cut))+
  geom_point() +
  geom_smooth()
```

geom_smooth has added a trendline 

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price, colour=cut))+
  geom_point() +
  geom_smooth()

print(obds_diamonds)
```


# Exercise 2

Predict the difference between these two plots

## Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

## Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer: Plot 2 has only one trendline, whereas plot 1 has several trendlines for each different cut. This is because colour=cut was included in the ggplot() command for plot 1, but not for plot 2

# Exercise 3

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight)) +
         geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet))+
  geom_point()
```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet))+
  geom_point()+
  geom_smooth()
```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer: diet 3 leads to the highest increase as this trendline is above the rest

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~Diet, nrow=2)
```

- Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickweight <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~Diet, nrow=2)
```

# Exercise 4

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
ggplot2::msleep
```

Type the name of the data-set to see a bit more about it

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(ggplot2::msleep, aes(x=order)) +
  geom_bar()
```

Note: must use geom_bar, not geom_histogram, as the data is not continuous

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ggplot2::msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1))
```
Use theme to change aspects of x axis text 
  axis.text.x must be specified to ensure that only the x axis text is affected
  element_text() is used to define angle, size, and hjust (justification, 1=right justified)


- Change the value and font size of the title for both x and y axes.

```{r}
ggplot(ggplot2::msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1)) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis")
```

Using theme can change the font size etc. for the axis titles, but labs must be used to change the words of the titles

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

```{r}
ggplot(ggplot2::msleep, aes(x=order, fill = genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1)) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis")
```

Adding fill=genus will colour the bars by genus, but makes the legend very large - need to fix this 

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ggplot2::msleep, aes(x=order, fill = genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1), legend.key.size = unit(2, "mm")) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis")
```

Using the command legend.key.size we can alter the size of each key in the legend - by setting unit(2,"mm") we have reduced the size

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ggplot2::msleep, aes(x=order, fill = genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1), legend.key.size = unit(2, "mm")) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis") +
  guides(fill=guide_legend(ncol=3))
```

Using the guides function, we set the number of columns in the legend to 3 

- Add a contour of thin black lines to the bars.

```{r}
ggplot(ggplot2::msleep, aes(x=order, fill = genus)) +
  geom_bar(colour="black", size=0.1) +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1), legend.key.size = unit(2, "mm")) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis") +
  guides(fill=guide_legend(ncol=3))
```
 
Adding colour="black" into the geom_bar function allows us to put a bar around each genus for clarity. Adding size=0.1 allows us to reduce the size of the line so that it is not too thick (default =1)
 
- Assign that last plot to an object called `obds_msleep`.

```{r}
obds_sleep <- ggplot(ggplot2::msleep, aes(x=order, fill = genus)) +
  geom_bar(colour="black", size=0.1) +
  theme(axis.text.x = element_text(angle=45, size=9, hjust = 1), legend.key.size = unit(2, "mm")) +
  labs(x="Taxonomic Order", y = "Number of Animals", title = "msleep dataset analysis") +
  guides(fill=guide_legend(ncol=3))
```

# Exercise 5

## Plotting grids

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

## Using `cowplot`.

Save each plot as its own row
Then use cowplot to combine each row into one plot

```{r}
library(cowplot)

first_row <- cowplot::plot_grid(obds_diamonds, labels="diamonds", ncol=1, nrow=1)
second_row <- cowplot::plot_grid(obds_chickweight, labels="chickweight", ncol=1, nrow=1)
third_row <- cowplot::plot_grid(obds_sleep, labels="sleep", ncol=1, nrow=1)
superplot<-plot_grid(first_row, second_row, third_row, ncol=1, nrow=3)

superplot
```

## Using `patchwork`.

Use / to place one plot on top of another plot 
Use + to place one plot next to another plot

```{r, fig.height=12, fig.width=12}
library(patchwork)

patchwork_plot <- obds_chickweight/obds_diamonds/obds_sleep
patchwork_plot

```

## Using `gridExtra`.

Similar to using cowplot

```{r, fig.height=12, fig.width=12}
library(gridExtra)

gridExtra_plot <- gridExtra::grid.arrange(obds_chickweight, obds_diamonds, obds_sleep, ncol=1, nrow=3)

gridExtra_plot
```

We can also try a different way of using the gridExtra software to get a different arrangement of graphs in the file

```{r, fig.height=12, fig.width=12}
library(gridExtra)

layout <- rbind(c(1,2), c(3,3))

gridExtra_plot_2 <- gridExtra::grid.arrange(obds_diamonds,obds_chickweight, obds_sleep, layout_matrix=layout)

```


## Exporting plots

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

We can use ggsave to save the various plots that we have created 
We give the plot name, the file extension, and define the width/height 

```{r}
ggsave("super_plot.pdf", plot=superplot, width=10, height=15)

ggsave("patchwork_plot.pdf", plot=patchwork_plot, width=10, height=15)

ggsave("gridExtra_plot.pdf", plot=gridExtra_plot, width=10, height=15)

```

# Exercise 6

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

geom_abline is used to draw a line on the plots 
    this is unrelated to the data, but just draws a slope - can be used for comparison)
    we have also added linetype (dashed) and size specifications in the function to define these aspects 

panel.grid.minor=element_blank is being used to remove the minor grid lines - this can help with clarity 

facet_wrap function is needed to draw an individual plot for each manufacturer 

labs has been used to add more informative x and y labels to the plot

strip.text=element_text in the theme function allows us to define the size of the text above each plot (i.e. the car manufacturer name) 

```{r}
ggplot(ggplot2::mpg, aes(x=cty, y=hwy, colour=year)) +
  geom_point(size=0.5) +
  geom_abline(intercept=0, slope=1, size=0.1, linetype="dashed")+
  facet_wrap(~ manufacturer) +
  theme(panel.grid.minor=element_blank(), strip.text=element_text(size=10)) +
  labs(y="Highway petrol useage", x="City petrol useage")
```