---
title: "single_cell_droplets"
author: "Hannah Pook"
date: '2022-05-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise 1

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)

# import raw counts
sce <- DropletUtils::read10xCounts(
  samples = c("raw" = "/project/obds/shared/resources/4_r_single_cell/pbmc_5k/raw_feature_bc_matrix/"))
  
# look at object   
sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
# calculate total UMI count for each sample 
colSums(counts(sce))

# find the position of cells with counts over 0
which(colSums(counts(sce)) >0)

# only keep barcodes where the total UMI count is greater than 0 (use [] to subset)
sce <- sce[,which(colSums(counts(sce)) >0)]

# look at object to check whether the dimensions make sense
sce
```

```{r}
# draw a graph to have a look at the remaining data (will help us decide lower for next function)

# first use barcodeRanks to extract count data 
rank_df <- DropletUtils::barcodeRanks(sce)

# use ggplot to plot data
ggplot(
  as.data.frame(rank_df),
  aes(x=rank, y=total)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_hline(yintercept = 100, colour = "turquoise")
```


- Run `DropletUtils::emptyDrops()`.

```{r}
# set seed for consistent results 
set.seed(123)

# run function with lower set to 100 (confirmed to be good by above plot)
out <- DropletUtils::emptyDrops(
  m = sce,
  lower = 100,
  niters = 10000
)

# have a look at the object
out

# see how many cells have UMIs over 100 
summary(out$Total>100)
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> The NA values represent cells which were not even tested - these are cells which had less than 100 UMIs 
> These cells were not tested as we set lower to 100 (i.e. we don't consider anything with <100 UMIs to be a cell, hence cells were automatically discarded)

- Have a look at the remaining data with a plot 
```{r}
ggplot(
  as.data.frame(out),
  aes(x=FDR)) +
  geom_histogram(fill = "turquoise") +
  theme_minimal()
```

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
# add a column to out to say whether cells are empty or not (is FDR <0.001)
out$empty <- out$FDR<0.001

# find position of non-empty cells in out
which(out$empty==TRUE)

# subset for non-empty cells
sce_filtered_emptydrops <- sce[,which(out$empty==TRUE)]
  

sce_filtered_emptydrops
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
# find mitochondrial genes 
is.mito <- grep("^MT-", rowData(sce)$Symbol)

library(scuttle)

# subset for mitochondrial genes 
sce <- scuttle::addPerCellQC(
  x = sce,
  percent_top = 50,
  subset = list(MT = is.mito))

# create column called empty in sce (says whether droplets are empty or not)
colData(sce)[["empty"]] <- out$empty

# create plots 
# note that scale = "width" is included only to adjust the scale of the graph 

# plot to show UMI values for empty and non-empty droplets 
plot1 <- colData(sce) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = empty, y = sum), scale = "width") +
    labs(x = "is cell containing droplet", y = "total UMI")

# plot to show features for empty and non-empty droplets
plot2 <- colData(sce) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = empty, y = detected), scale = "width") +
    labs(x = "is cell containing droplet", y = "features")

# plot to show % mito for empty and non-empty droplets 
plot3 <- colData(sce) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = empty, y = subsets_MT_percent), scale = "width") +
    labs(x = "is cell containing droplet", y = "MT percent")

cowplot::plot_grid(plot1, plot2, plot3, nrow = 1)
  
```

We can see: 
- The UMI count is higher in non-empty droplets 
- The feature count is higher in non-empty droplets 
- Empty droplets contain some mitochondrial RNA (this is likely to occur by chance)

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  


all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise 2

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
# set seed for consistency
set.seed(123)

# run doublet finder
sce_scdblfinder <- scDblFinder::scDblFinder(
  sce = sce_filtered_emptydrops,
  clusters = TRUE)

# see columns in new object 
colData(sce_scdblfinder)

# make a table of the number of singlets vs doublets 
table(colData(sce_scdblfinder)$scDblFinder.class)

# plot a heatmap of the doublets found between different clusters 
plotDoubletMap(sce_scdblfinder)
```
```{r}
# run the function again without clustering
sce_scdblfinder_v2 <- scDblFinder::scDblFinder(
  sce = sce_filtered_emptydrops,
  clusters = FALSE)

# make a table of the number of singlets vs doublets 
table(colData(sce_scdblfinder_v2)$scDblFinder.class)

# now only have 6 columns
colData(sce_scdblfinder_v2)

# this function will no longer work as there are no clusters
plotDoubletMap(sce_scdblfinder_v2)
```


**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
# find which cells are singlets
singlets_index <- which(colData(sce_scdblfinder)$scDblFinder.class=="singlet")

# filter for singlets 
sce_filtered_scdblfinder <- sce_scdblfinder[,singlets_index]
  
# view new object
sce_filtered_scdblfinder

metadata(sce_filtered_scdblfinder)$scDblFinder.stats
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
# find mitochondrial genes 
is.mito <- grep("^MT-", rowData(sce_scdblfinder)$Symbol)

# subset based on mitochondrial genes 
library(scuttle)

sce_scdblfinder <- scuttle::addPerCellQC(
  x = sce_scdblfinder,
  subset = list(MT = is.mito))

# view new object 
colData(sce_scdblfinder)

# create plots 

# plot to show UMI values for singlets and doublets 
plotA <- colData(sce_scdblfinder) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = scDblFinder.class, y = sum), scale = "width") +
    labs(x = "cell class", y = "total UMI")

# plot to show features for singlets and doublets
plotB <- colData(sce_scdblfinder) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = scDblFinder.class, y = detected), scale = "width") +
    labs(x = "cell class", y = "features")

# plot to show % mito for singlets and doublets
plotC <- colData(sce_scdblfinder) %>%
    as_tibble() %>% 
    ggplot() +
    geom_violin(aes(x = scDblFinder.class, y = subsets_MT_percent), scale = "width") +
    labs(x = "cell class", y = "MT percent")

cowplot::plot_grid(plotA, plotB, plotC, nrow = 1)
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise 3

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}
# we have added z = to give the clusters we calculated previously with scDblFinder 
sce_decontX <- celda::decontX(
  x = sce_filtered_scdblfinder,
  z = sce_filtered_scdblfinder$scDblFinder.cluster)

# here we have not added z 
sce_decontX_v2 <- celda::decontX(
  x = sce_filtered_scdblfinder, 
  varGenes = 2000)
```

- Create histogram of contamination column
```{r}
# plot the first sce_decontX
ggplot(
  as.data.frame(colData(sce_decontX)),
  aes(x = decontX_contamination)) +
  geom_histogram(fill = "lightslateblue") +
  theme_minimal()

# plot the second sce_decontX
ggplot(
  as.data.frame(colData(sce_decontX_v2)),
  aes(x = decontX_contamination)) +
  geom_histogram(fill = "aquamarine") +
  theme_minimal()
```

This plot shows us that most cells have a low level of contamination 

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
# check reduced dimensions available 
reducedDimNames(sce_decontX)

# this shows that the only reduced dims matrix available is decontX_UMAP - decontX has done its own UMAP  


# version 1 code 
# we need to extract UMAP co-ordinates for the first version
umap <- reducedDim(sce_decontX, "decontX_UMAP")
  
# plot clusters for first version of sce_decontX
plotDimReduceCluster(
  x = sce_decontX$decontX_clusters, 
  dim1 = umap[,1],
  dim2 = umap[,2]
)

# version 2 code 
# we need to extract UMAP co-ordinates for the second version
umap_v2 <- reducedDim(sce_decontX_v2, "decontX_UMAP")

# plot clusters for second version of sce_decontX
plotDimReduceCluster(
  x = sce_decontX_v2$decontX_clusters, 
  dim1 = umap_v2[,1],
  dim2 = umap_v2[,2]
)

# second version looks better so use this downstream
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX_v2)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> Some clusters seem to have higher levels of contamination
> Comparing to the previous plot, cluster 8 seems to have highest levels of contamination 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}
# replace the rownames (currently ENSEMBL ID) with gene symbols or keep ENSEMBL ID if the name is not unique 
# ID acts as a backup to names 
rownames(sce_decontX_v2) <- uniquifyFeatureNames(
  ID = rowData(sce_decontX_v2)$ID,
  names = rowData(sce_decontX_v2)$Symbol)

# calculate log normalised counts (better for plotting)
sce_decontX_v2 <- logNormCounts(
  x = sce_decontX_v2,
  assay.type = "decontXcounts")

# create feature plot, coloured based on expression of given genes 
plotDimReduceFeature(
  x = assay(sce_decontX_v2, "logcounts"),
  features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"),
  dim1 = umap_v2[,1],
  dim2 = umap_v2[,2]
  )
  
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
# create a list of marker genes 
markers <- list(
  Tcell_Markers = c("CD3E", "CD3D"), 
  Bcell_Markers = c("CD79A", "CD79B", "MS4A1"), 
  Monocyte_Markers = c("S100A8", "S100A9", "LYZ"), 
  NKcell_Markers = "GNLY"
)

# create a list of which cluster corresponds to which cell type 
cellTypeMappings <- list(
  "T_cells" = c(1,3),
  "B_cells" = c(4),
  "Monocytes" = c(2,9),
  "NK_cells" = c(6)
)

# plot marker genes before and after decontamination 
# will automatically use both "counts" and "decontXcounts" assays
plotDecontXMarkerPercentage(
  x = sce_decontX_v2,
  markers = markers,
  groupClusters = cellTypeMappings)
```
This plot shows us that the cell identities correlate better with expected markers following decontamination with decontX method 


```{r}
# can add assayName if only want to plot one assay e.g. just decontaminated counts 
plotDecontXMarkerPercentage(
  x = sce_decontX_v2,
  markers = markers,
  groupClusters = cellTypeMappings, 
  assayName = "decontXcounts")
```


- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r}
plotDecontXMarkerExpression(
  x = sce_decontX_v2,
  markers = markers,
  groupClusters = cellTypeMappings
  ) +
  theme_minimal()
```

We can see that after the decontamination, expression of inappropriate/unexpected markers for different cell types decreases
